---
title: "R Notebook"
output: html_notebook
---

```{r setup}
library(car)
```


**Variable Descriptions:**

```{r load}
load("bwght_w203.RData")
desc
```

**Exploratory Analysis of each variable:**

- mage

```{r mage}
hist(data$mage)
summary(data$mage)
```

Nothing out of ordinary. Normal distribution. No transformation required

- meduc

```{r meduc}
hist(data$meduc)
summary(data$meduc)
```

Atypical distribution, with spikes probably due to education milestones. Probably could transform to indicator or categorical variable.

- monpre

```{r monpre}
hist(data$monpre)
summary(data$monpre)
```

What does zero mean? Should we take it out?

- npvis 

```{r npvis}
hist(data$npvis, breaks = 40)
summary(data$npvis)
```

Somewhat normal. Peak at around 11 (monthly visits?)

- fage

```{r fage}
hist(data$fage, breaks = 40)
summary(data$fage)
```

Nothing out of ordinary. Normal distribution, slight positive skew. No transformation required

- meduc

```{r feduc}
hist(data$feduc)
summary(data$feduc)
```

Atypical distribution, with spikes probably due to education milestones. Probably could transform to indicator or categorical variable.

- bwght

```{r bwght}
hist(data$bwght, breaks = 50)
summary(data$bwght)
```

Normal, negative skew. No transformation required, but log-transform could make practical sense.

- omaps

```{r omaps}
hist(data$omaps)
summary(data$omaps)
```

One to ten scale, not continuous, does not make much sense.

-fmaps

```{r fmaps}
hist(data$fmaps)
summary(data$fmaps)
```

One to ten scale, not continuous, does not make much sense.

- cigs

```{r cigs}
hist(data$cigs, breaks = 40)
hist(data$cigs[data$cigs > 0], breaks = 40)
summary(data$cigs)
```

Weird distribution, with a few spikes at marked points due to survey answer round-ups or round-downs. Lots and lots of zeroes. Perhaps change to indicator or categorial variable?

- drink

```{r drink}
hist(data$drink)
hist(data$drink[data$drink > 0], breaks = 40)
summary(data$drink)
```

Lots and lots of zeroes, just like cigs. Perhaps transform to indicator or categorical variable.

- male

```{r male}
hist(data$male)
summary(data$male)
```

Approximately equal sample distribution.

- male

```{r mwhte}
hist(data$mwhte)
summary(data$mwhte)
```

Uneven, mostly white mothers.

Verify correlation of every variable:

```{r scatterplotmatrix}
scatterplotMatrix(~ bwght + mage + meduc + monpre + npvis + fage + feduc + cigs + drink, data=data)
scatterplotMatrix(~ bwght + omaps + fmaps + male + mwhte + mblck + moth + fwhte + fblck + foth, data=data)
```